---
title: "Bloomington Animal Care and Control Data"
---
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Bloomington Animal Care and Control data is available through **[BClear](https://data.bloomington.in.gov/)**, a data portal for Bloomington, Indiana's city government. The **[Animal Care and Control Division](https://bloomington.in.gov/animal-shelter)** has prepared two data sets that the public may use to understand how the city handles animal welfare. 

Documention and processing workflow for two Bloomington Animal Care and Control data sets are provided below. These data sets include the division's historic **Annual Statistics** and **Animal Cases Data** representing the flow of animals in and out of the Bloomington Animal Shelter from the start of 2017 to present day. 

This notebook was created for the **BMG Hack's - Let's Get Visual - Civic Data Challenge**. The notebook provides a description of each data set, and provides code for the data processing performed on these data that (hopefully) will facilitate analysis and visualization that can be updated, re-used, and shared with others! 

### Download Processed Data
Data processing outputs are collected in a ZIP file and available for download in the link below. Data is provided in CSV format.

#### **[Processed Animal Care and Control  Data](https://github.com/mginda/bmghack-animalcontrol/raw/master/data/bmg-acc-data.zip)** - Feb. 20, 2020

### Navigate to...

* [Annual Statistics Documentation](#annual-stats)
* [Animal Case Data Documentation](#cases)

# Data Documentation and Processing Notes

## R Environment Set-Up

Here we set up the R environment, and load in the R Packages that will be used in the notebook.

```{r enviro, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
#Clean up environment
rm(list=ls())
#Packages used in notebook
library(curl)        # Data downloads via URL
library(tidyr)       # Tidyverse data processing piping, string manipulation
library(readr)       # Data parsing
library(plyr)        # Combining and Splitting Data
library(dplyr)       # Data analysis and aggregations
library(lubridate)   # Dates
library(magrittr)    # Piping
```

### Loading Data from Animal Care and Control

Data is loaded directly into R using a CURL function that uses **BClear** URLS to data downloads for the Animal Care and Control data. Data is provided a CSVs with headers.

```{r data-load, echo=TRUE}
#Annual Statistics
annual <- read.csv(curl(url="https://data.bloomington.in.gov/dataset/9a0fc126-0775-4952-96ff-76ece67ffbc7/resource/7a847ec2-31c6-48e3-b0bc-02af2fa94587/download/animal-care-and-control-data-set-2004-2019.csv"), header=TRUE, stringsAsFactors = F)
#Intake and adoption records
intake <- read.csv(curl(url="https://data.bloomington.in.gov/dataset/94d3f457-57b5-45be-bee0-a0106f59b7ed/resource/8854ce02-e8f5-44b9-b85f-17f002a7d023/download/animals.csv"), header=TRUE, stringsAsFactors = F)
```

Before analyzing and visualizing these data, the variables in the data set were reviewed to ensure that they have the correct data type in R, and to determine how variables relate to each other. The data processing routine documented below is used to clean-up and reshape the data to aid development and resuse of these data in a visual analytics workflow.

## Annual Statistics Documentation {#annual-stats}

#### **Source: [Annual Statistics (2004-2019)](https://data.bloomington.in.gov/dataset/animal-care-and-control/resource/7a847ec2-31c6-48e3-b0bc-02af2fa94587)** 

These data represent high level annual statistics for the division, which includes revenue, budgets, staffing, counts of animal intakes and movements. These data is great for those interested in practicing implementing basic data visualizations using a tidy data set.

* Number of employees - *employees* 
* Number of vehicles - *vehicles*
* Annual Budget - *budget* - $
* Grants Recieved - *grants* - $
* Revenue from Adoptions - *revenue* - $
* Service Calls - *serviceCalls*
* Emergency Calls - *emergencyCalls*
* Number of Owner Surrended - *surrender*
* Number of Strays - *stray*
* Number of Impounded - *impound*
* Total Number of Intakes - *intake*
* Number of Transfer to other Shelters - *transfer*
* Number of Fosters - *foster*
* Number of Adoptions - *adopt*
* Number of Returnes - *return*
* Number of Euthinizations - *euthenize*

### Data Structure

A review of the data structure reveals two issues with the Animial Care and Control Divisions' Annual Statistics data set:

1. Variables use verbose names with have had punctuation added for spaces and text is captialized;
2. Monetary values are captured as character strings with symbols and empty spaces rather than providing providing a numeric values. We need to fix these two issues with our data.

```{r str-an, echo=FALSE}
str(annual[,c(4,16,17)])
```

### Preparing Data

```{r pro-an, echo = T, results = 'hide'}
#Save original column name values
annual_cols <- colnames(annual)
#Substitutes periods from original column names with spaces
annual_cols <- gsub("\\.","\\ ",annual_cols)
#Renames fields to shorter values
names(annual) <- c("year","employees","vehicles","budget","surrender","stray","impound",
                   "intake","adopt","return","euthanize","tranfer","foster","serviceCalls",
                   "emergencyCalls","grants","revenue")
#Parse the budget, grants and revenue fields to create numeric values instead of character strings
annual$budget <- parse_number(annual$budget, na="NA")
annual$grants <- parse_number(annual$grants, na="NA")
annual$revenue <- parse_number(annual$revenue, na="NA")
```

Now we can get accurate descriptive data summaries for the `r length(colnames(annual))` variables in the Annual Statistics observations for the Animal Control Division. Variables measured, include:

### Summary Descriptive Statistics

```{r summary-an, echo=FALSE}
#Annual Statistics
summary(annual)
```

### Example Annual Statistics Observations

```{r head-an, echo=FALSE}
#Annual Statistics
head(annual)
```

## Animal Case Data Documenation {#cases}

#### **Source: [Adoptions (2017-Present)](https://data.bloomington.in.gov/dataset/adopted-animals/resource/8854ce02-e8f5-44b9-b85f-17f002a7d023)** 

The Animal Case data set represent animal case records maintained by the division. This is a much larger and complex data set compared to the annual statistics. The Bloomington Animal Shelter, which uses the [Shelter Manager](https://www.sheltermanager.com/site/en_tour.html) database to manage its case load. The data provided by the city is a roll-up of multiple tables from the shelter database. Currently, there are over 12,000 intake records available to download.

The goal of data processing workflow presented below is to create different subsets that best represent the different types of interactions that an animal may experience in the shelter system. These sub-sets allow a user to create tailored analysis and visualizations. The data sets derived from the origianl data set include:

1. [Animal Identification](#animals)
2. [Intake Cases](#intakes)
3. [Animals on the Move (adoptions, fosters, transfers, etc.)](#moves)
4. [Returned Animals](#returns)
5. [Deceased Animals](#dead)
6. [Shelter Interaction Log](#log)
7. [Supplemental Animal Statistics](#sup)

### Known Issues and Limitations 

Working with the Animal Case Data set has provided me some insights on issues and limitations present in these data. Issues relate to the quality of data observations captured by the shelter, while limitations indicate areas where there are gaps in what is knowable based on the data provided.

#### Issues found in data
* Animal ages are estimates of the intake specialists based on examination of the animal and/or from information provided upon intake & return.
* Animal breed information may be the best guess of the intake specialists.
* Reasons for animal intake and returns are subjective and limited to one category per animal. Some assignments may be defaults for particular cases where multiple reasons apply.
* Transfers to other shelters exist within intake and movement data sets; however, the flag field *istransfer* is not used in these interactions.
* The calculation of duration in the shelter system is subjective. I consider time in foster care to be time an animal is still in the system, but outside of the shelter. 
* All animals are assigned a reason/cause for being killed; however, this is likely a default field required at intake, even if an animal has not been killed. These values are excluded from the *dead* data set.
* Some animals taken or returned are logged as *dead on arrival*/*DOA*, but also have a duration time calculation.
* Flags used in the *dead* data set have unclear combinations (e.g. an animal died off shelter and is DOA). 

#### Limitations with data
* There is not data on who brought in, adopted, reclaimed, or fostered an animal.
* Transfers to other shelters exist within intake and movement data sets; however, where an animal comes or goes is not provided here.
* The outcome of an animal transferred to another shelter system is unknown.
* Some adopted animals may have been returned to another shelter, or given to a family or friend against shelter adoption agreements.
* Animal build/size is captured but not provided here.
* Animal fertility status is not capture in these data.
* Animal special needs and behavioral issues may not be fully captured.
* Animal sponsorship for adoption by staff and volunteers is not captured in the data.
* The shelter runs multiple adoption focused events; however, these events are not provided as a data set to supplement case records.

### Processing Workflow

#### Data Structure

There are `r length(rownames(intake))` observations across `r length(colnames(intake))` variables in this data set. The data structure provides a description of the variables below. 

Reading the variable names shows that the data set uses common naming convention in variable names to indicate relationships. Examples of variable relationships include fields starting with: *intake*, *movement*, *return*, and *deceased*. This suggests that there are a minimum of four tables combined in this data set. 

Data tables are likely joined in the shelter's database by the *id* acting as the primary key  This field will present in all derived subsets to help join related data frames.

```{r str-in, echo=FALSE}
#Intake Structure
str(intake)
```

#### Assigning Data Types
Before subsetting the animal case data set into cleaner subsets, variables are processed to ensure that they have the proper data types assigned. The case data requires transforming fields into factor/categorical variables, and to dates. Note, *animalage* is ignored, but it is processed later in the *Animal Idenfification* data set.

```{r pro-in-datatypes, echo=TRUE, warning=TRUE}
#Copy intake to new data.frame data 
data <- intake
#List of factor variables column names
intake_fac <- colnames(data[,c(1,3,5:10,12,13,15,18,20)])
#Converts variables from character strings to factors
for(i in 1:length(intake_fac)){
  data[,intake_fac[i]] <- data[,intake_fac[i]] %>% as.factor()
}
#List of date variable column names
intake_date <- colnames(data[,c(2,14,17,19)])
#Convert variables from character strings to dates
for(i in 1:length(intake_date)){
  if(length(data[data[,intake_date[i]]=="",intake_date[i]])>0){
    data[data[,intake_date[i]]=="",intake_date[i]] <- NA
  }
  data[,intake_date[i]] <- data[,intake_date[i]] %>% as.Date()
}
rm(i, intake_date, intake_fac)
```

#### 1. Animal Identification {#animals}

Animals taken in by Animal Care and Control Division are described across a variety of features, including multiple identifiers, a basic description of the animal, and the location within the shelter. The variables included in this data set are:

* Animal Case identifier - *id* - a numeric unique identifier with `r length(unique(intake$id))` unique values.
  * This field acts as the primary key that will connect data sets together
* Shelter Code - *sheltercode* - a numeric unique identifier with `r length(unique(intake$sheltercode))`values.
  * The difference between unique identifers and total number of observations in the data set (`r length(rownames(intake))-length(unique(intake$sheltercode))`) indicates that there may be multiple intake records for a given animal in the data set.
  * The letter code starting each id indicates species class (e.g. `r intake[1,]$sheltercode` for cat).
* Animal identity chip - *identichipnumber* - factor variable with `r length(unique(intake$identichipnumber))` unique values
  * Not all animals have chip identifiers.
* Animal name - *animalname* - factor variable with `r length(unique(intake$animalname))` unique values
  * We've got some common names!
* Breed - *breedname* - factor variable with `r length(unique(intake$breedname))` unique values
* Animal coloration - *basecolour* - factor variable with `r length(unique(intake$basecolour))` unique values
* Species - *speciesname* - factor variable with `r length(unique(intake$speciesname))` unique values.
  * There are some surprises here
* Animal age (estimated) - *animalage* - string with values such as `r intake[1,]$animalage`
  * This field is valuable, but requires processing to extract a usable estimate of age (years or months)
* Animal sex - *sex* - factor variable with `r length(unique(intake$sexname))` unique values.

```{r pro-animals, echo=TRUE, warning=TRUE}
#Creates demographic data set of unique animals with intake cases
animals <- data[,c("id","sheltercode","identichipnumber","animalname","animalage","sexname","speciesname","breedname","basecolour")] %>% unique()
#Processes animal age to
tmp <- as.data.frame(as.character(animals$animalage),stringsAsFactors = F)
names(tmp) <- "animalage"
tmp$intakeage <- as.numeric(0)
for(i in 1:nrow(tmp)){
  #Converts string to a dataframe by way of 2 col matrice
  d <- stringr::str_split(tmp[i,]$animalage,"\\ ") %>% unlist() 
  if(length(d)==2){
    d <- d %>% matrix(ncol=2) %>% data.frame()
  } else {
    d <- d %>% matrix(ncol=2) %>% t() %>% data.frame()
  }
  d$X1 <- as.character(d$X1) %>% as.numeric()
  # Years to months
  if(length(d[d$X2=="years" | d$X2=="year",]$X2)>0){
    d[d$X2=="years" | d$X2=="year",]$X1 <- d[d$X2=="years" | d$X2=="year",]$X1 * 12
  }
  #Weeks to months
  if(length(d[d$X2=="weeks." | d$X2=="week.",]$X2)>0){
    d[d$X2=="weeks." | d$X2=="week.",]$X1 <- d[d$X2=="weeks." | d$X2=="week.",]$X1 * (1/4.5)
  }
  #Days to months
  if(length(d[d$X2=="days." | d$X2=="day.",]$X2)>0){
    d[d$X2=="weeks." | d$X2=="week.",]$X1 <- d[d$X2=="weeks." | d$X2=="week.",]$X1 * (1/30)
  }
  tmp[i,]$intakeage <- sum(d$X1)
}
#Creates new column in animals data.frame
animals$intakeage <- tmp$intakeage
#Removes objects
rm(i,d,tmp)
#Summarize intake cases
summary(animals)
```

### 2. Intake Cases {#intakes}

Intake Case data include any variable that relates to when an animal first interacts with the Animal Care and Control shelter system. Intake related variables included in this data set are:

* Intake identifier - *id* - a numeric unique identifier with `r length(unique(intake$id))` unique values.
  * The number of unqiue ids and unique shelter codes are identical. And, as with *sheltercodes* variable above, the count of unique *id* values is less than total intake records provided by the division. This suggests unique identifier is linked to the animal's intial intake.
* Intake date - *intakedate* - changed to -> *date* - a date field (YYYY-MM-DD) ranging from `r min(as.Date(intake$intakedate))` to `r max(as.Date(intake$intakedate))`.
  * Intake data sets may require more examination and potentiall filtering to removing intakes prior 2017, the first year when records begin were first collected in full.
  * Needs processing to as.Date() type.
* Intake reason - *intakereason*  - changed to -> *interaction* - a factor variable with `r length(unique(intake$intakereason))` unique values.
* Location within shelter - *location* - a factor variable with `r length(unique(intake$location))` unique values.
* Interation type - *type* - a new field that indicates the type of interaction with the shelter

```{r pro-intake, echo=TRUE, warning=TRUE}
#Creates demographic data set of unique animals with intake cases
intakes <- data[,c("id","intakedate","intakereason","location")] %>% unique()
#Update field names
names(intakes)[2:3] <- c("date","interaction")
#Updates blank level with None
levels(intakes$interaction)[levels(intakes$interaction)==""] <- "No Reason"
#Interaction type
intakes$type <- "intake"
#Summarize intake cases
summary(intakes)
```

### 3. Animals on the Move {#moves}

After the initial intake, the goal of the shelter is to move animals out of the Animal Care and Control system. Some example movements include when an animal being fostered or adopted. Variables related to animal movements, include:

* Animal Case identifier - *id* - a numeric unique identifier with `r length(unique(intake$id))` unique values.
* Movement date - *movementdate* - changed to -> *date* - a date field (YYYY-MM-DD) ranging from `r min(as.Date(intake$movementdate))` to `r max(as.Date(intake$movementdate))`.
  * Needs processing to as.Date() data types.
* Movement type - *movementtype* - changed to -> *interaction* - a factor variable with `r length(unique(intake$movementtype))` unique values.
  * This field indicates what happens to an animal after the intake is logged..
* Trial status - *istrial* - a binary (1/0) variable indicating if the animal is a trial adoption.
  * All values are currently *0*, meaning the field has no value and should be removed.
* Transfer status - *istransfer* - a binary (1/0) variable indicating if the animal is transferred from the system. 
* Interation type - *type* - a new field that indicates the type of interaction with the shelter system.

```{r pro-move, echo=TRUE, warning=TRUE}
#Creates demographic data set of unique animals with intake cases
move <- data[,c("id","movementdate","movementtype")] %>% 
  unique() %>% 
  arrange(id)
names(move)[2:3] <- c("date","interaction")
move$type <- "move"
#Summarize movements
summary(move)
```

### 4. Returned Animals {#returns}

Occasionally, an animal returns to the shelter. However, rather than creating a new intake case, the animal's interaction is logged as a return. Variables related to returned to the shelter system describe cause and date for a return to Animal Care and Control. The variables included in this data set are:

* Animal Case identifier - *id* - a numeric unique identifier with `r length(unique(intake$id))` unique values.
* Return date - *returndate* - changed to -> *date* - a date field (YYYY-MM-DD) ranging from `r min(as.Date(intake[intake$returndate!="",]$returndate))` to `r max(as.Date(intake[intake$returndate!="",]$returndate))`.
  * Not all intake records include a return date, indicating the animal has not been returned to the system.
  * Needs processing to as.Date() data types.
* Return reason - *returnedreason* - changed to -> *interaction* - a factor variable with `r length(unique(intake$returnedreason))` unique values.
  * Some interesting unique values include: `r head(unique(intake$returnedreason),3)`. 
  * This variable may be an interesting feature to explore reasons based on animal type and demographics. This may be useful for a predictive model at intake.
* Interation type - *type* - a new field that indicates the type of interaction with the shelter system.
  
```{r pro-return, echo=TRUE, warning=TRUE}
#Creates demographic data set of unique animals with intake cases
return <- data[,c("id","returndate","movementdate","returnedreason")] %>%
  unique() %>% 
  arrange(id)
return <- return[!is.na(return$returndate),]
#Update field names
names(return)[c(2,4)] <- c("date","interaction")
#Add interaction type
return$type <- "return"
#Time to return
return$duration <- difftime(return$date,return$date, unit="day")
#Summarize intake cases
summary(return)
```

### 5. Deceased Animals {#dead}

Sometimes animals taken in or returned to the shelter that cannot be released. In these cases, the Animal Care and Control Division will euthanize an animal, or in rare cases, document a death that takes place at intake. The variables that describe data related to animal that have died while managed by Animal Care and Control included in this data set are: 

* Animal Case identifier - *id* - a numeric unique identifier with `r length(unique(intake$id))` unique values.
* Deceased date - *deceaseddate* - changed to -> *date* - a date field (YYYY-MM-DD) ranging from `r min(as.Date(intake[intake$deceaseddate!="",]$deceaseddate))` to `r max(as.Date(intake[intake$deceaseddate!="",]$deceaseddate))`.
  * Not all intake records include a deceased date, indicating the animal is alive!
  * Needs processing to as.Date() data types.
* Deceased reason - *deceasedreason* - changed to -> *interaction* - a factor variable with `r length(unique(intake$deceasedreason))` unique values.
  * Some interesting unique values include: `r head(unique(intake$deceasedreason),3)`. 
  * This variable may be an interesting feature to explore reasons based on animal type and demographics. This may be useful for a predictive model at intake.
  * All animals have a deceased reason even if they are not dead.
* Animal died off selter - *diedoffshelter* -  a binary (1/0) variable indicating if the animal is died outside of the shelter.
* Euthanize flag - *puttosleep* - a binary (1/0) variable indicating if the animal is euthanized from the system.
* Dead on arrival flag - *isdoa* - a binary (1/0) variable indicating if the animal is died when they arrived at the shelter.
* Interation type - *type* - a new field that indicates the type of interaction with the shelter system.

```{r pro-doa, echo=TRUE, warning=TRUE}
#Creates demographic data set of unique animals with intake cases
dead <- data[,c("id","deceaseddate","deceasedreason",
                "diedoffshelter","puttosleep","isdoa")] %>% unique()
dead <- dead[!is.na(dead$deceaseddate),]
#Update field names
names(dead)[2:3] <- c("date","interaction")
#Adds interaction type
dead$type <- "dead"
#Summarize intake cases
summary(dead)
```

### 6. Shelter Interaction Log {#log}

The Shelter Interaction Log provides a time series data set provides another data structure that may be relevant for a variety of visual analytic and modeling workflows. This data set captures the intake and outcomes data for each case managed by the shelter as a list. The list is ordered by the *Case ID* and *Date* of an interactions with the shelter system. This data may be joined to the animal descriptive data to help subset or filter the interaction logs.

* Animal Case identifier - *id* - a numeric unique identifier with `r length(unique(intake$id))` unique values.
* Interaction - *interaction* - a factor variable that describes the interaction with the shelter system. These categories have duplicate and similar values that are updated to reduce the quantity of qualitiative values.
* Interation type - *type* - a factor variable that indicates the type of interaction with the shelter system (e.g intake, movement, return, dead).
* Interaction sequence - *seq* - Number represents an animal's sequential interaction with the shelter system.
* Start date - *date* - a date field (YYYY-MM-DD) for when an interaction with the shelter system begins.
* End date - *date_end* - a date field (YYYY-MM-DD) for when an interaction with the shelter system ends.
* Time Interval - *time_int* - an interval field indicating period of time when an interaction occured.
* Duration - *duration* - Measure of time between an interaction with the shelter system, measured in days.

```{r pro-ani-log}
#Creating shelter interaction dataframe
interact <- rbind(intakes[,c("id","date","interaction","type")],
                    move[,c("id","date","interaction","type")],
                    return[,c("id","date","interaction","type")],
                    dead[,c("id","date","interaction","type")]) %>% 
                arrange(id,date)

#Intraction categories updated
tmp <- interact$interaction %>% as.character() 
#change interaction to... Behaviors 
tmp[tmp %in% c("Return Adopt - Behavior","Temperament/Behavior")] <- "Behavior"
#change interaction to... Dead on arrival
tmp[tmp %in% c("DOA","Cremation")] <- "Dead On Arrival"
#change interaction to... UU-untreatable, unmanageable Behavior
tmp[tmp == c("UU-untreatable, unmanageable Behavior")] <- "UU - untreatable, unmanageable Behavior"
#change interaction to animal health
tmp[tmp %in% c("Return Adopt - Animal Health","Sick/Injured","Medical")] <- "Animal Health"
#change interaction to owner deceased
tmp[tmp %in% c("Owner Died")] <- "Owner Deceased"
#change interaction to... Owner requested
tmp[tmp %in% c("Owner requested Euthanasia")] <- "Owner requested"
#change interaction to... Incompatible with owner lifestyle
tmp[tmp %in% c("Return adopt - lifestyle issue")] <- "Incompatible with owner lifestyle"
tmp <- as.factor(tmp)
#Add cleaned interaction data to data table
interact$interaction <- tmp

#Loop calculates duration between shelter interaction for each animal's case
interact$date_end <- NA
interact$seq <- 0
for(i in 1:length(unique(intakes$id))){
  tmp <- interact[interact$id==animals[i,]$id,c("id","date","type","date_end","seq")]
  if(nrow(tmp)>1){
        tmp[1:nrow(tmp)-1,]$date_end <- tmp[2:nrow(tmp),]$date
  }
  #Introduction of the NA in the date_end field are treated
  if(nrow(tmp[is.na(tmp$date_end) & tmp$type=="dead",])==1){
    #In cases where the type of interaction is a death, the NA date_end field is the same       the date field
    tmp[is.na(tmp$date_end),]$date_end <- tmp[is.na(tmp$date_end),]$date
  } else {
    #In cases where the type of interaction is an intake, return, or move, date_end is the      current system date
    tmp[is.na(tmp$date_end),]$date_end <- Sys.Date()
  }
  #Interaction sequence
  tmp$seq <- seq_along(tmp$id)
  
  #Adds back tmp back to 
  interact[interact$id==intakes[i,]$id,c("date_end","seq")] <- tmp[,c("date_end","seq")]
  }
#Clean environment
rm(tmp,i)

#Format values for date_end field
interact$date_end <-as.Date(interact$date_end, origin=lubridate::origin,
                            format="%Y-%m-%d")

#Checks date_end for DOA on return
interact[interact$type=="return" & interact$interaction=="Dead On Arrival",]$date_end <-
  interact[interact$type=="return" & interact$interaction=="Dead On Arrival",]$date

#Create time interval that interaction takes place
interact$time_int <- interval(interact$date,interact$date_end)
#Calculate difference in time (days) of interval
interact$duration <- difftime(interact$date_end,interact$date,unit="days")

#Transforms type into a factor variables
interact <- interact[,c(1,3,4,6,2,5,7:8)]

#Summary of data
summary(interact)
```

### 7. Supplemental Statistic for Animals {#sup}

Supplemental statistics for animals provide additional measurements that describe an animal's time and interaction in the shelter. These variables were calculated from the **Shelter Interaction Log**. The list of supplemental statistical measured are described below. The data manipulation scripts to calculate supplemental statistics are documented in the preceeding sections.

* Number of Interactions with Shelter - *interactions* - this numeric value indicates the number of interactions that an animal has with the shelter. 
* Total time in shelter system - *timeinsys* - this numeric value representing the number of days that an animal is in the Animal Care and Control Division Shelter system.
* Transfer In - *transfer_in* - a binary flag field indicating is an animal transfered into the shelter.
* Transfer Out - *transfer_out* - a binary flag field indicating is an animal transfered out of the shelter.
* Adopted - *adopt* & *adopt_ct* - a binary flag field and count field indicating animal was adopted.
* Fostered - *foster* & *foster_ct* - a binary flag field and count field indicating animal was fostered.
* Reclaimed - *reclaim* & *reclaim_ct* - a binary flag field and count field indicating animal was reclaimed by an owner. 
* Returned - *return* & *return_ct* - a binary flag field and count field indicating animal was returned to the shelter. 
* Death Logged by Shelter - *dead* - a binary flag field and count field indicating animal has died by shelter staff.
* Shelter Intake Date - *intakedate* - date an animal was first taken in by the shelter.

#### Supplemental Statistics Data Frame

A copy of the **Animal Idenification** data set is copied into a new data frame.

```{r a-sup}
#New data frame of supplemental stats
animals_s <- animals
```

#### Number of Interactions with Shelter

The **Number of Interactions with Shelter** uses the **Shelter Interaction Log** data. First, the *id* and *interaction* variables selected from the shelter's interaction log (*interact*), then aggregated by the *group_by()* funtion on an animal's identifier. Last, the the number interactions is counted using the *summarise()* and *n()* functions.

```{r a-sup-interactions}
#Total Shelter Interactions (Intake+Returns+Movements+Deaths)
tmp <- interact %>% 
          select(id,interaction) %>% 
          group_by(id) %>% 
          summarise(interactions = n()) 
animals_s <- join(animals_s,tmp,by="id")
```

#### Time in the Shelter System

**Time in the Shelter System** value represents the sum total duration in days that an animal is in the shelter or being fostered. Periods of time that are apart of this calculation include time from intake to a move out of the system (e.g. time to adoption, released to wilderness, or time until reclaimed by an owner.) Periods of time an animal moves out of the system, but later returns (e.g. adopted, reclaimed by owner, escaped, or stolen but returned to the shelter) are excluded from the calculations.

For the calculation, the **Shelter Interaction Log** is filtered to remove duration measurements with *NA* values are removed (e.g typically, last interaction with the shelter, such as when an animal has been adopted or when an animal is put down.)), and to remove a set of movement interactions with duration values associated with time outside of the shelter's system (e.g. when an animal isn't fostered). Next, the *id* and *duration* fields are selected, and aggregated by the *group_by()* funtion on an animal's identifier. Finally, then total duration calculated using the  *summarise()* and *sum()* functions.

```{r a-sup-timesys}
#Time in Selter System
#Included durations: period between intakes or returns until a movement or death; time in foster care until movement or death;
#Exclude durations: periods where an animal was adopted, reclaimed or stolen. Foster time 
tmp <- interact %>% 
          filter(!is.na(duration) & 
                 !(interaction %in% c("Adoption", "Reclaimed", "Stolen",
                                      "Escaped", "Released to Wilderness"))) %>% 
          select(id,duration) %>% 
          group_by(id) %>% 
          summarise(timeinsys = sum(duration)) 
animals_s <- join(animals_s,tmp,by="id")
```

#### Adoption, Foster and Reclaimed Animal Flags and Counts

The following processes calculate a number of fields related to common animal movements. First, a logic field is calculated to indicate if an animal was transfered out of the shelter, adopted, fostered, and or reclaimed by the animal's owner. Next, the count of times an animal is adopted, fostered or reclaimed are calculated. These movement may appear multiple times in the shelter interaction log.

```{r a-sup-moves}
#Flags animals by type of common movement
tmp <- interact %>% 
        filter(type=="move") %>%
        group_by(id) %>%
        summarise(
          transferout = any(interaction=="Transfer"),
          adopt = any(interaction=="Adoption"),
          foster = any(interaction=="Foster"),
          reclaim = any(interaction=="Reclaimed")
        ) 
#T/F to 1/0 
tmp[,2:5] <- lapply(tmp[,2:5], as.numeric)
#Counts of common movements
tmp$reclaim_ct <- tmp$foster_ct <- tmp$adopt_ct <-  0
#Number of time animal was Adopted
tmp2 <- interact %>% 
          filter(type=="move" & interaction=="Adoption") %>%
          group_by(id) %>%
          summarise(ct = n())
tmp[tmp$id %in% tmp2$id,]$adopt_ct <- tmp2$ct
#Count of times animal fostered
tmp2 <- interact %>% 
          filter(type=="move" & interaction=="Foster") %>%
          group_by(id) %>%
          summarise(ct = n())
tmp[tmp$id %in% tmp2$id,]$foster_ct <- tmp2$ct
#Count of times animal reclaimed
tmp2 <- interact %>% 
          filter(type=="move" & interaction=="Reclaimed") %>%
          group_by(id) %>%
          summarise(ct = n())
tmp[tmp$id %in% tmp2$id,]$reclaim_ct <- tmp2$ct
#Joins tmp file to supplemental data frame
animals_s <- join(animals_s,tmp,by="id")
#Re-order columns
animals_s <- animals_s[,c(1:14,17,15,18,16,19)]
rm(tmp, tmp2)
```

#### Return Flag and Counts

The following processes calculates a number of fields animals returned to the system. First, a logic field is calculated to indicate that an animal returned to the shelter system. Next, the count of times an animal was returned is calculated, as an animal may return multiple times.

```{r a-sup-returns}
#Flags animals by type of common movement
tmp <- interact %>% 
        filter(type=="return") %>%
        group_by(id) %>%
        summarise(return=any(type=="return"),
                  return_ct=n())
#T/F to 1/0 
tmp[,2] <- lapply(tmp[,2], as.numeric)
#Join fields to supplemental stat data frame
animals_s <- join(animals_s,tmp,by="id")
#Updates NA fields to 0 values for animals that have not been returned
animals_s[is.na(animals_s$return),c(20:21)] <- 0
```

#### Transfer Into Shelter Flag

```{r ani-sup-tin}
tmp <- interact %>% 
        filter(type=="intake") %>%
        group_by(id) %>%
        summarise(
          transferin = any(interaction=="Transfer from Other Shelter"),
        )
#T/F to 1/0 
tmp[,2] <- lapply(tmp[,2], as.numeric)
#Join field to supplemental stats
animals_s <- join(animals_s,tmp,by="id")
#Reorder fields
animals_s <- animals_s[,c(1:12,22,13:21)]
```

#### Death Logged by Shelter Flag

```{r ani-sup-dead}
tmp <- dead %>% 
        group_by(id) %>%
        summarise(
          dead = 1)
#Join field to supplemental stats
animals_s <- join(animals_s,tmp,by="id")
#Reorder fields
animals_s[is.na(animals_s$dead),]$dead <- 0
```

#### Shelter Intake Date

```{r ani-sup-indate}
tmp <- interact %>% 
        group_by(id) %>%
        summarise(intakedate = first(date))
# Add field to supplemental stats
animals_s <- join(animals_s,tmp,by="id")
rm(tmp)
```